\plaintitle

### 2020-03-22 wip-design/paper.tex
Working on implementation section

\section{Implementation}
\label{sec:implementation}

\subsection{Language}

\subsection{Data cube}

\subsection{Tabular data}

\subsection{Graph database}

@@

\begin{table}
  \centering
  \begin{tabular}{l l l c l}
    \toprule
      & {\small \textit{Task}}
      & {\small \textit{Kind}} & {\small \textit{Done}}
      & {\small \textit{Notes}} \\
    \midrule
    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
    \small \#13 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
    \bottomrule
  \end{tabular}
  \caption{Overview of the work completed by individual participants. The marks denote:
    \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
  \label{tab:tasks}
  \vspace{-1em}
\end{table}

@@

\paragraph{Benefits and drawbacks of text}

AG - confused by indentation (Monaco inserts it automatically - duh!)
it does not work if you have it and add a new to-level thing;

BM uses copy & paste to copy thing and replace one identifier in the middle
(which would be difficult to do with non-textual representation)

people occasionally miss '.' when typing and try to use a space

\paragraph{How users understand the then keyword}

what is then: "allows us to chain together the operations"
is it special? "maybe I need it just to split over multiple lines" (experiments..) "no!"

"when you explained about the "dot then" that was a really useful thing to know. When I found that, I was like 'this is fine, this is doable'. If I knew this from the start, it would [have been easier]" (~27)

"I thought 'then' starts a new line. (...) I realised there was no 'then' after the 'paging' command. "
"It starts a new line or a new query? I assumed there would be one after 'paging' too."

@@

\subsection{Conclusions}
(former journalist)
Do you think this is something journalists could figure out how to use?
"Yeah, I think so. There's a lot of effort going into data journalism that programming could make much quicker,
but I was always nervous about code." ... "Something like this would really simplify things."

>>>>

\section{Implementation}
\label{sec:implementation}
The Gamma consists of a programming language with a web-based coding environment and a number
of type providers that provide access to various kinds of data sources. The system is implemented
in F\# and compiled to JavaScript using Fable.

\subsection{The Gamma Language}
In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
or an expression that evaluates to a value such as a data table or a chart.
An expression is a reference to a data source followed by a chain of member accesses.
A member can be either an ordinary member or an operation which takes a list of parameters
enclosed in parentheses. Our implementation supports a few other features, but those
are not used in this paper.

The language uses a type system to infer what members are available at a given point in a chain
of members. The type is an object with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by a type provider for a particular data source.
The types are also generated lazily. A type is only computed when the corresponding member
is actually used in code.

The web-based coding environment uses the Monaco\footnote{\url{https://microsoft.github.io/monaco-editor/}}
editor and implemts auto-completion feature that is triggered when the user types `.' and relies
on information from the type system. The can display the value of a current command inline
as in Figure~\ref{fig:thegamma}, or values of all commands on the side as in Figure~\ref{fig:sidebyside}.
In the latter case, it can also combine one or more data series into a single chart.

\subsection{Data Cube Type Provider}

\subsection{Tabular Data Type Provider}

\subsection{Graph Database Type Provider}

\newpage

\begin{figure}
\includegraphics[width=1\columnwidth]{figures/sidebyside}
\caption{The Gamma environment with an automatically generated chart comparing CO2 emissions
of China and United States.}
\label{fig:sidebyside}
\vspace{-0.5em}
\end{figure}

@@

\begin{table}
  \centering
  \begin{tabular}{l l l c l}
    \toprule
      & {\small \textit{Task}}
      & {\small \textit{Kind}} & {\small \textit{Done}}
      & {\small \textit{Notes}} \\
    \midrule
    \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one of two data series}\\
    \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
    \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
    \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
    \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
    \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
    \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
    \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
    \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
    \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
    \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
    \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
    \bottomrule
  \end{tabular}
  \caption{Overview of the work completed by individual participants. The marks denote:
    \priorityc{75} = required some guidance, \priorityc{50} = partially completed.}
  \label{tab:tasks}
  \vspace{-1em}
\end{table}

@@

\paragraph{Benefits and drawbacks of text}
The Gamma is based on text to aid transparency. Text hints that there is no hidden state and
the reader sees the full code. The study suggests that using a text editor has both
benefits and drawbacks compared to alternatives such as structured editors~\cite{structure-based,livenut,lamdu}.
Most participants had no difficulty navigating around source code, making edits or deleting code
fragments, which is arguably harder in an unfamiliar structured editor.

We observed two issues in the study. Participant \#2 struggled with correct indentation, starting a second command
with more indentation than needed and participant \#6 had a syntax error in an unrelated command,
which prevents charts from rendering. Some participants used the text editor effectively,
e.g.~participant \#5, who used copy-and-paste to fetch the same data series for multiple countries.

@@

\paragraph{How users understand the `then' member}
When interviewing participants who worked with a tabular data source, we
asked about their understanding of the \ikvd{then} member. This is a regular member generated by the
type provider (i.e.~not a keyword), but it has a special meaning. Consider
the following example which calculates the average travel expenses and total number of
representatives per county using the UK House of Lords expenses data:

\begin{thegamma}
expneses.'group data'.'by County'
  .'average Travel Costs'.'count distinct Name'.then
\end{thegamma}

The \ikvd{then} member is used to complete a part of a query where the user can continue
adding items to build a list. Here, we select two aggregations to be calculated for each
county before choosing \ikvd{then} and applying other operations such as sorting.

Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
over multiple lines, but rejected the idea after experimenting and noting that they can insert
line breaks elsewhere. One correctly concluded that it ``allows us to chain together the
operations'' of the query. Participant \#13 needed a hint about using the
\ikvd{then} member and later reflected:

\begin{quote}
  \emph{``When you explained about the `dot then' that was a really useful thing to know.
  When I found that, I was like this is fine, this is doable. If I knew this from the start,
  it would [have been easier].''}
\end{quote}

This comment summarizes an important fact. Although iterative prompting allows the users to start
exploring new data sources, the domain specific languages used by more complex data sources have
their own design principles that users need to understand to use the data source effectively.

\subsection{Conclusions}
The motivation behind The Gamma is to make simple programmatic data exploration accessible to
journalists and other non-experts. Main-stream programming languages have the benefit of
reproducibility and transparency, but are difficult to use. Our study shows that non-experts
are able to solve basic data exploration tasks using The Gamma. When asked whether The Gamma
is something that journalists could learn how to use, participant \#13 (a former journalist) answered:

\begin{quote}
\emph{``Yeah, I think so. There's a lot of effort going into data journalism that
  programming could make much quicker, but I was always nervous about code. (...)
  Something like this would really simplify things.''}
\end{quote}

The Gamma is based on the iterative prompting principle which lowers the barrier to entry, but
it is worth noting that it does not fully eliminate complexity involved in data querying.
Instead, it provides a way of structuring the complexity. Iterative prompting makes it easy to
get started, but complex data sources will inevitably require additional learning. The Gamma
makes this easier by allowing users to learn from published code samples.
\newpage
~
\newpage
